---
title: "01_data_import_and_QC"
author: "Melanie Smith"
date: "2026-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects includuing hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = FALSE, dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reprodcibility
```

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)

# RNA-seq
library(edgeR)
library(limma)
```

# Set paths

```{r}
projectDir <- "/home/melanie-smith/workDir/natashaHarvey/2026-shanna_hlec_FAT4_kd_flow_response"
umiDir <- file.path(projectDir, "RNASeq", "raw_data", "umi_processing") # UMI pipeline outputs copied from the HPC
countsFile <- file.path(umiDir, "featurecounts", "featurecounts_dedup.txt")
multiqcDir <- file.path(umiDir, "multiqc_umi", "multiqc_data")
starDir <- file.path(umiDir, "star_align_umi")
outDir <- file.path(projectDir, "RNASeq", "out_dir")
gtfFile <- file.path(projectDir, "RNASeq", "genomeFiles", "gencode.v39.annotation.gtf")

dir.create(outDir, showWarnings = FALSE) # out_dir was archived on 2026-10-07; recreate it if missing
```

# Import data

UMI-deduplicated gene counts from featureCounts (`-p --countReadPairs -s 2`, see `RNASeq/hpc_scripts/scripts/`). One row per gene (versioned Ensembl ID) with annotation columns (`Chr`, `Start`, `End`, `Strand`, `Length`), then one column per library named by the full path of its deduplicated BAM. Counts are fragments (read pairs).

```{r}
fcCounts <- read_tsv(countsFile, comment = "#", show_col_types = FALSE)

# simplify column names to the SAGC ULN, e.g. ".../umi_dedup/26-03807_S1_L03.dedup.bam" -> "26-03807_S1_L03"
colnames(fcCounts) <- str_replace(colnames(fcCounts), "^.*/(26-\\d{5}_S\\d+_L03)\\.dedup\\.bam$", "\\1")

dim(fcCounts)
stopifnot(n_distinct(fcCounts$Geneid) == nrow(fcCounts)) # one row per gene

rawCountMat <- fcCounts %>%
  dplyr::select(Geneid, starts_with("26-")) %>%
  column_to_rownames("Geneid") %>%
  as.matrix()

dim(rawCountMat)
stopifnot(ncol(rawCountMat) == 13) # 12 samples + SAGC negative control
stopifnot(!anyNA(rawCountMat), all(rawCountMat >= 0)) # complete, non-negative counts
```

# Pipeline QC summary

Per-library metrics for each step of the UMI pipeline, from the MultiQC tables: UMI extraction pass rate, read pairs kept after trimming, STAR mapping, UMI deduplication (`dedupKept` = % of mapped reads that are unique molecules, i.e. library complexity) and deduplicated fragments assigned to genes by featureCounts.

```{r}
ulnPattern <- "26-\\d{5}_S\\d+_L03" # SAGC ULN, used to match sample names across the MultiQC tables

pipelineQC <- read_tsv(file.path(multiqcDir, "multiqc_umitools_extract.txt"), show_col_types = FALSE) %>%
  transmute(ULN         = str_extract(Sample, ulnPattern),
            inputPairs  = extract_input_reads,
            extractPass = extract_percent_passing) %>%
  left_join(read_tsv(file.path(multiqcDir, "multiqc_cutadapt.txt"), show_col_types = FALSE) %>%
              transmute(ULN          = str_extract(Sample, ulnPattern),
                        trimmedPairs = pairs_written),
            by = "ULN") %>%
  left_join(read_tsv(file.path(multiqcDir, "multiqc_star.txt"), show_col_types = FALSE) %>%
              transmute(ULN            = str_extract(Sample, ulnPattern),
                        uniquelyMapped = uniquely_mapped_percent,
                        multiMapped    = multimapped_percent),
            by = "ULN") %>%
  left_join(read_tsv(file.path(multiqcDir, "multiqc_umitools_dedup.txt"), show_col_types = FALSE) %>%
              transmute(ULN           = str_extract(Sample, ulnPattern),
                        dedupKept     = dedup_percent_passing,
                        meanUMIperPos = mean_umi_per_pos),
            by = "ULN") %>%
  left_join(read_tsv(paste0(countsFile, ".summary"), show_col_types = FALSE) %>%
              filter(Status == "Assigned") %>%
              pivot_longer(-Status, names_to = "Sample", values_to = "assignedFragments") %>%
              transmute(ULN = str_extract(Sample, ulnPattern), assignedFragments),
            by = "ULN")

stopifnot(nrow(pipelineQC) == 13, !anyNA(pipelineQC)) # one complete row per library
pipelineQC
```

# Strandedness

STAR `ReadsPerGene.out.tab` from the alignment of the UMI-extracted, trimmed reads (before deduplication), one per library in `star_align_umi/<ULN>/`. The first four rows are STAR summary counts (`N_unmapped`, `N_multimapping`, `N_noFeature`, `N_ambiguous`); the remaining rows are genes. Columns are: gene ID, unstranded counts, forward-stranded counts (read 1 on the gene strand), and reverse-stranded counts (read 1 on the opposite strand). The library prep is reverse-stranded, so nearly all gene-assigned reads should be in the reverse column. Check each library using the forward fraction, forward / (forward + reverse), and keep the reverse-stranded counts as the trimmed, not deduplicated, count matrix for comparison with the deduplicated counts.

```{r}
ulnIDs <- list.files(starDir) # one sub-directory per library, named by the SAGC ULN
stopifnot(length(ulnIDs) == 13)

readsPerGene <- file.path(starDir, ulnIDs, "ReadsPerGene.out.tab") %>%
  set_names(ulnIDs) %>%
  map(~ read_tsv(.x, col_names = c("GeneID", "unstranded", "forward", "reverse"), col_types = "ciii"))

geneIDs <- readsPerGene[[1]]$GeneID
stopifnot(all(map_lgl(readsPerGene, ~ identical(.x$GeneID, geneIDs)))) # same genes in the same order in every file
stopifnot(identical(head(geneIDs, 4), c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))) # STAR summary rows first

readsPerGeneLong <- bind_rows(readsPerGene, .id = "ULN")

strandedness <- readsPerGeneLong %>%
  filter(!str_starts(GeneID, "N_")) %>%
  group_by(ULN) %>%
  summarise(unstranded = sum(unstranded),
            forward    = sum(forward),
            reverse    = sum(reverse)) %>%
  mutate(fracForward = forward / (forward + reverse))

strandedness
stopifnot(all(strandedness$fracForward[strandedness$ULN != "26-03819_S13_L03"] < 0.1)) # all real libraries reverse-stranded

trimCountMat <- readsPerGeneLong %>%
  filter(!str_starts(GeneID, "N_")) %>%
  dplyr::select(ULN, GeneID, reverse) %>% # reverse-stranded counts (column 4)
  pivot_wider(names_from = ULN, values_from = reverse) %>%
  column_to_rownames("GeneID") %>%
  as.matrix()

stopifnot(setequal(rownames(trimCountMat), rownames(rawCountMat)), setequal(colnames(trimCountMat), colnames(rawCountMat))) # same genes and libraries as the deduplicated counts
trimCountMat <- trimCountMat[rownames(rawCountMat), colnames(rawCountMat)] # same order as the deduplicated counts

dim(trimCountMat)
stopifnot(!anyNA(trimCountMat), all(trimCountMat >= 0)) # complete, non-negative counts
```

# QC of the SAGC negative control

Library `26-03819_S13_L03` is the SAGC negative control (library conc. <0.1 ng/uL, 2,690 read pairs; the 31.7M clusters passing filter in the SAGC PDF report is an error). Compare the number of deduplicated fragments assigned to genes and the number of genes detected against the real libraries before removing it.

```{r}
tibble(ULN           = colnames(rawCountMat),
       assignedReads = colSums(rawCountMat),
       genesDetected = colSums(rawCountMat > 0))
```

# Drop the negative control and rename samples

Map the SAGC ULN to the sample IDs used in the protein analysis (SAGC quality report SAGCQR2280). The negative control is not in the key, so it is dropped here.

```{r}
sampleKey <- tribble(
  ~ULN,               ~sampleID,
  "26-03807_S1_L03",  "CSMT16", # CS16
  "26-03808_S2_L03",  "FSMT16", # FS16
  "26-03809_S3_L03",  "CLMT16", # CL16
  "26-03810_S4_L03",  "FLMT16", # FL16
  "26-03811_S5_L03",  "CSMT17", # CS17
  "26-03812_S6_L03",  "FSMT17", # FS17
  "26-03813_S7_L03",  "CLMT17", # CL17
  "26-03814_S8_L03",  "FLMT17", # FL17
  "26-03815_S9_L03",  "CSMT18", # CS18
  "26-03816_S10_L03", "FSMT18", # FS18
  "26-03817_S11_L03", "CLMT18", # CL18
  "26-03818_S12_L03", "FLMT18"  # FL18
)

countMat <- rawCountMat[, sampleKey$ULN]
colnames(countMat) <- sampleKey$sampleID

trimCountMat <- trimCountMat[, sampleKey$ULN] # trimmed, not deduplicated, for comparison
colnames(trimCountMat) <- sampleKey$sampleID

dim(countMat)
stopifnot(ncol(countMat) == 12, !"26-03819_S13_L03" %in% colnames(countMat)) # 12 samples, negative control removed
```

# Sample metadata

Genotype and condition are encoded in the first two letters of the sample ID (C/F = Control/FAT4 KO; L/S = Laminar/Static).

```{r}
sampleMeta <- tibble(sampleID = colnames(countMat)) %>%
  mutate(genotype  = factor(if_else(str_sub(sampleID, 1, 1) == "C", "Control", "FAT4 KO"), levels = c("Control", "FAT4 KO")),
         condition = factor(if_else(str_sub(sampleID, 2, 2) == "L", "Laminar", "Static"), levels = c("Static", "Laminar")),
         group     = paste(genotype, condition, sep = "_"))

sampleMeta

saveRDS(countMat, file.path(outDir, "01_count_matrix.rds"))
saveRDS(sampleMeta, file.path(outDir, "01_sample_metadata.rds"))
```

# Gene annotation

Gene-level annotation from the GENCODE v39 GTF (the annotation used for STAR alignment). Keep only the gene records and extract the gene name, biotype and location for each versioned Ensembl ID. The table is saved to `out_dir` so later scripts don't need to re-read the GTF.

```{r}
geneAnno <- read_tsv(gtfFile, comment = "#", col_names = c("chr", "source", "type", "start", "end", "score", "strand", "frame", "attributes"),
                     col_types = "cccddcccc") %>%
  filter(type == "gene") %>%
  mutate(GeneID   = str_extract(attributes, '(?<=gene_id ")[^"]+'),
         geneName = str_extract(attributes, '(?<=gene_name ")[^"]+'),
         geneType = str_extract(attributes, '(?<=gene_type ")[^"]+')) %>%
  dplyr::select(GeneID, geneName, geneType, chr, start, end, strand)

dim(geneAnno)
stopifnot(n_distinct(geneAnno$GeneID) == nrow(geneAnno)) # one row per gene
stopifnot(setequal(geneAnno$GeneID, rownames(countMat))) # annotation matches the count matrix exactly

geneAnno <- geneAnno[match(rownames(countMat), geneAnno$GeneID), ] # same order as the count matrix
count(geneAnno, geneType, sort = TRUE)

saveRDS(geneAnno, file.path(outDir, "01_gene_annotation.rds"))
```

# Library sizes

Library size here is the total number of UMI-deduplicated fragments assigned to genes in the count matrix (not the clusters passing filter in the SAGC report).

```{r}
libSize <- tibble(sampleID = colnames(countMat),
                  libSize  = colSums(countMat)) %>%
  left_join(sampleMeta, by = "sampleID") %>%
  arrange(genotype, condition) %>%
  mutate(sampleID = factor(sampleID, levels = sampleID)) # order samples by group on the x-axis

libSize
```

```{r, fig.width=7, fig.height=5}
libSizePlot <- ggplot(libSize, aes(x = sampleID, y = libSize / 1e6, fill = group)) +
  geom_col() +
  scale_fill_manual(values = c("Control_Static"   = "#6BAED6",
                               "Control_Laminar"  = "#08519C",
                               "FAT4 KO_Static"   = "#FD8D3C",
                               "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = NULL,
       y = "Library size (million deduplicated fragments)",
       title = "Library sizes (UMI-deduplicated)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

libSizePlot
ggsave(file.path(outDir, "01_library_sizes.png"), libSizePlot, width = 7, height = 5, dpi = 300)
```

# Library size and gene detection before and after deduplication

Compare the trimmed, not deduplicated, STAR counts with the UMI-deduplicated featureCounts counts. `pctRetained` is the % of gene-assigned fragments that are unique molecules. STAR GeneCounts and featureCounts use slightly different assignment rules, so the comparison is approximate. The per-library pipeline QC, with sample IDs, is saved for the GC content vs library complexity comparison in `02_filter_normalise_MDS.Rmd`.

```{r}
dedupComparison <- tibble(sampleID       = colnames(countMat),
                          libSizeTrimmed = colSums(trimCountMat),
                          libSizeDedup   = colSums(countMat),
                          genesTrimmed   = colSums(trimCountMat > 0),
                          genesDedup     = colSums(countMat > 0)) %>%
  mutate(pctRetained = 100 * libSizeDedup / libSizeTrimmed) %>%
  left_join(sampleMeta, by = "sampleID") %>%
  arrange(genotype, condition)

dedupComparison

pipelineQC %>%
  left_join(sampleKey, by = "ULN") %>% # sampleID is NA for the negative control
  write_csv(file.path(outDir, "01_pipeline_QC.csv"))
```

```{r, fig.width=8, fig.height=5}
dedupPlot <- dedupComparison %>%
  mutate(sampleID = factor(sampleID, levels = sampleID)) %>% # order samples by group on the x-axis
  dplyr::select(sampleID, libSizeTrimmed, libSizeDedup) %>%
  pivot_longer(-sampleID, names_to = "stage", values_to = "libSize") %>%
  mutate(stage = factor(stage, levels = c("libSizeTrimmed", "libSizeDedup"), labels = c("Before deduplication", "After deduplication"))) %>%
  ggplot(aes(x = sampleID, y = libSize / 1e6, fill = stage)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Before deduplication" = "grey70",
                               "After deduplication"  = "grey20")) +
  labs(x = NULL,
       y = "Library size (million fragments assigned to genes)",
       fill = NULL,
       title = "Library sizes before and after UMI deduplication") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

dedupPlot
ggsave(file.path(outDir, "01_library_sizes_before_after_dedup.png"), dedupPlot, width = 8, height = 5, dpi = 300)
```

# MDS of raw data

MDS of the unfiltered, unnormalised UMI-deduplicated counts on the log2 CPM scale. `plotMDS` uses the top 500 genes by leading log2 fold change between each pair of samples.

```{r}
logCPM <- edgeR::cpm(countMat, log = TRUE)

mds <- limma::plotMDS(logCPM, plot = FALSE)
varExplained <- round(100 * mds$var.explained, 1)

mdsDf <- tibble(sampleID = colnames(logCPM),
                dim1     = mds$x,
                dim2     = mds$y) %>%
  left_join(sampleMeta, by = "sampleID")
```

```{r, fig.width=7, fig.height=5}
mdsPlot <- ggplot(mdsDf, aes(x = dim1, y = dim2, colour = group, shape = condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = sampleID), vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Control_Static"   = "#6BAED6",
                                 "Control_Laminar"  = "#08519C",
                                 "FAT4 KO_Static"   = "#FD8D3C",
                                 "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = paste0("Leading logFC dim 1 (", varExplained[1], "%)"),
       y = paste0("Leading logFC dim 2 (", varExplained[2], "%)"),
       title = "MDS of raw log2 CPM (UMI-deduplicated)") +
  theme_bw()

mdsPlot
ggsave(file.path(outDir, "01_MDS_raw_logCPM.png"),
       mdsPlot,
       width = 7,
       height = 5,
       dpi = 300)
```

# Session information

```{r}
sessionInfo()
```
