---
title: "01_data_import_and_QC"
author: "Melanie Smith"
date: "2026-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects includuing hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = FALSE, dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reprodcibility
```

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)

# RNA-seq
library(edgeR)
library(limma)
```

# Set paths

```{r}
projectDir <- "/home/melanie-smith/workDir/natashaHarvey/2026-shanna_hlec_FAT4_kd_flow_response"
starDir <- file.path(projectDir, "RNASeq", "raw_data", "star_align")
```

# Import data

STAR `ReadsPerGene.out.tab` files, one per library in `star_align/<ULN>/`. The first four rows are STAR summary counts (`N_unmapped`, `N_multimapping`, `N_noFeature`, `N_ambiguous`); the remaining rows are genes. Columns are: gene ID, unstranded counts, forward-stranded counts (read 1 on the gene strand), and reverse-stranded counts (read 1 on the opposite strand).

```{r}
ulnIDs <- list.files(starDir) # one sub-directory per library, named by the SAGC ULN
stopifnot(length(ulnIDs) == 13)

readsPerGene <- file.path(starDir, ulnIDs, "ReadsPerGene.out.tab") %>%
  set_names(ulnIDs) %>%
  map(~ read_tsv(.x, col_names = c("GeneID", "unstranded", "forward", "reverse"), col_types = "ciii"))

geneIDs <- readsPerGene[[1]]$GeneID
stopifnot(all(map_lgl(readsPerGene, ~ identical(.x$GeneID, geneIDs)))) # same genes in the same order in every file
stopifnot(identical(head(geneIDs, 4), c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))) # STAR summary rows first

readsPerGeneLong <- bind_rows(readsPerGene, .id = "ULN")
```

# STAR alignment summary

Key alignment metrics from each library's STAR `Log.final.out`.

```{r}
file.path(starDir, ulnIDs, "Log.final.out") %>%
  set_names(ulnIDs) %>%
  map(~ read_delim(.x, delim = "|", col_names = c("metric", "value"), trim_ws = TRUE, show_col_types = FALSE)) %>%
  bind_rows(.id = "ULN") %>%
  filter(metric %in% c("Number of input reads", "Uniquely mapped reads %",
                       "% of reads mapped to multiple loci", "% of reads unmapped: too short")) %>%
  pivot_wider(names_from = metric, values_from = value)
```

# STAR assignment summary

Per-library STAR summary counts (reverse-stranded column) alongside the number of reads assigned to genes.

```{r}
readsPerGeneLong %>%
  dplyr::select(ULN, GeneID, reverse) %>%
  mutate(GeneID = if_else(str_starts(GeneID, "N_"), GeneID, "assigned")) %>%
  group_by(ULN, GeneID) %>%
  summarise(reads = sum(reverse)) %>%
  pivot_wider(names_from = GeneID, values_from = reads)
```

# Strandedness

The library prep is reverse-stranded, so nearly all gene-assigned reads should be in the reverse column. Check each library using the forward fraction, forward / (forward + reverse), then build the raw count matrix from the reverse-stranded counts.

```{r}
strandedness <- readsPerGeneLong %>%
  filter(!str_starts(GeneID, "N_")) %>%
  group_by(ULN) %>%
  summarise(unstranded = sum(unstranded),
            forward    = sum(forward),
            reverse    = sum(reverse)) %>%
  mutate(fracForward = forward / (forward + reverse))

strandedness
stopifnot(all(strandedness$fracForward[strandedness$ULN != "26-03819_S13_L03"] < 0.1)) # all real libraries reverse-stranded

rawCountMat <- readsPerGeneLong %>%
  filter(!str_starts(GeneID, "N_")) %>%
  dplyr::select(ULN, GeneID, reverse) %>% # reverse-stranded counts (column 4)
  pivot_wider(names_from = ULN, values_from = reverse) %>%
  column_to_rownames("GeneID") %>%
  as.matrix()

dim(rawCountMat)
stopifnot(!anyNA(rawCountMat), all(rawCountMat >= 0)) # complete, non-negative counts
```

# QC of the SAGC negative control

Library `26-03819_S13_L03` is the SAGC negative control (library conc. <0.1 ng/uL, but 31.7M clusters passing filter). Compare the number of reads assigned to genes and the number of genes detected against the real libraries before removing it.

```{r}
tibble(ULN           = colnames(rawCountMat),
       assignedReads = colSums(rawCountMat),
       genesDetected = colSums(rawCountMat > 0))
```

# Drop the negative control and rename samples

Map the SAGC ULN to the sample IDs used in the protein analysis (SAGC quality report SAGCQR2280). The negative control is not in the key, so it is dropped here.

```{r}
sampleKey <- tribble(
  ~ULN,               ~sampleID,
  "26-03807_S1_L03",  "CSMT16", # CS16
  "26-03808_S2_L03",  "FSMT16", # FS16
  "26-03809_S3_L03",  "CLMT16", # CL16
  "26-03810_S4_L03",  "FLMT16", # FL16
  "26-03811_S5_L03",  "CSMT17", # CS17
  "26-03812_S6_L03",  "FSMT17", # FS17
  "26-03813_S7_L03",  "CLMT17", # CL17
  "26-03814_S8_L03",  "FLMT17", # FL17
  "26-03815_S9_L03",  "CSMT18", # CS18
  "26-03816_S10_L03", "FSMT18", # FS18
  "26-03817_S11_L03", "CLMT18", # CL18
  "26-03818_S12_L03", "FLMT18"  # FL18
)

countMat <- rawCountMat[, sampleKey$ULN]
colnames(countMat) <- sampleKey$sampleID

dim(countMat)
stopifnot(ncol(countMat) == 12, !"26-03819_S13_L03" %in% colnames(countMat)) # 12 samples, negative control removed
```

# Sample metadata

Genotype and condition are encoded in the first two letters of the sample ID (C/F = Control/FAT4 KO; L/S = Laminar/Static).

```{r}
sampleMeta <- tibble(sampleID = colnames(countMat)) %>%
  mutate(genotype  = factor(if_else(str_sub(sampleID, 1, 1) == "C", "Control", "FAT4 KO"), levels = c("Control", "FAT4 KO")),
         condition = factor(if_else(str_sub(sampleID, 2, 2) == "L", "Laminar", "Static"), levels = c("Static", "Laminar")),
         group     = paste(genotype, condition, sep = "_"))

sampleMeta
```

# Library sizes

Library size here is the total number of reads assigned to genes in the count matrix (not the clusters passing filter in the SAGC report).

```{r}
libSize <- tibble(sampleID = colnames(countMat),
                  libSize  = colSums(countMat)) %>%
  left_join(sampleMeta, by = "sampleID") %>%
  arrange(genotype, condition) %>%
  mutate(sampleID = factor(sampleID, levels = sampleID)) # order samples by group on the x-axis

libSize
```

```{r, fig.width=7, fig.height=5}
ggplot(libSize, aes(x = sampleID, y = libSize / 1e6, fill = group)) +
  geom_col() +
  scale_fill_manual(values = c("Control_Static"   = "#6BAED6",
                               "Control_Laminar"  = "#08519C",
                               "FAT4 KO_Static"   = "#FD8D3C",
                               "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = NULL,
       y = "Library size (million assigned reads)",
       title = "Library sizes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# MDS of raw data

MDS of the unfiltered, unnormalised counts on the log2 CPM scale. `plotMDS` uses the top 500 genes by leading log2 fold change between each pair of samples.

```{r}
logCPM <- edgeR::cpm(countMat, log = TRUE)

mds <- limma::plotMDS(logCPM, plot = FALSE)
varExplained <- round(100 * mds$var.explained, 1)

mdsDf <- tibble(sampleID = colnames(logCPM),
                dim1     = mds$x,
                dim2     = mds$y) %>%
  left_join(sampleMeta, by = "sampleID")
```

```{r, fig.width=7, fig.height=5}
ggplot(mdsDf, aes(x = dim1, y = dim2, colour = group, shape = condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = sampleID), vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Control_Static"   = "#6BAED6",
                                 "Control_Laminar"  = "#08519C",
                                 "FAT4 KO_Static"   = "#FD8D3C",
                                 "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = paste0("Leading logFC dim 1 (", varExplained[1], "%)"),
       y = paste0("Leading logFC dim 2 (", varExplained[2], "%)"),
       title = "MDS of raw log2 CPM") +
  theme_bw()
```

# Session information

```{r}
sessionInfo()
```
