---
title: "01_data_import_and_PCA"
author: "Melanie Smith"
date: "2026-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects includuing hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = FALSE, dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr’s “grouped output” informational messages
set.seed(42) # set seed for reprodcibility
```

# Load Libraries

```{r}
# Core tidyverse
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Set paths

```{r}
projectDir <- "/home/melanie-smith/workDir/natashaHarvey/2026-shanna_hlec_FAT4_kd_flow_response"
inFile <- file.path(projectDir, "Protein", "raw_data", "After_Normalisation_imputation.csv")
```

# Import data

The input file is precursor-level (one row per peptide precursor), so the protein group quantities are repeated across rows. Keep only the protein identifiers and the 12 `PG.Quantity` columns, then collapse to one row per protein group.

```{r}
rawData <- read_csv(inFile, show_col_types = FALSE)

pgQuant <- rawData %>%
  dplyr::select(PG.ProteinGroups, PG.Genes, ends_with(".PG.Quantity")) %>%
  distinct()

# simplify column names to the sample ID, e.g. "[1] CLMT16_S2-C1_1_3471.d.PG.Quantity" -> "CLMT16"
colnames(pgQuant) <- str_replace(colnames(pgQuant), "^\\[\\d+\\] ([A-Z]+\\d+)_S2-.*$", "\\1")

dim(pgQuant)
stopifnot(n_distinct(pgQuant$PG.ProteinGroups) == nrow(pgQuant)) # one row per protein group
```

# Sample metadata

Genotype and condition are encoded in the first two letters of the sample ID (C/F = Control/FAT4 KO; L/S = Laminar/Static).

```{r}
sampleMeta <- tibble(sampleID = colnames(pgQuant)[-(1:2)]) %>%
  mutate(genotype  = factor(if_else(str_sub(sampleID, 1, 1) == "C", "Control", "FAT4 KO"), levels = c("Control", "FAT4 KO")),
         condition = factor(if_else(str_sub(sampleID, 2, 2) == "L", "Laminar", "Static"), levels = c("Static", "Laminar")),
         group     = paste(genotype, condition, sep = "_"))

sampleMeta
```

# PCA

PG.Quantity values are on the linear scale, so log2 transform before PCA.

```{r}
log2Mat <- pgQuant %>%
  column_to_rownames("PG.ProteinGroups") %>%
  dplyr::select(all_of(sampleMeta$sampleID)) %>%
  as.matrix() %>%
  log2()

# samples as rows, proteins as columns
pca <- prcomp(t(log2Mat), center = TRUE, scale. = FALSE)
varExplained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pcaDf <- as.data.frame(pca$x[, 1:4]) %>%
  rownames_to_column("sampleID") %>%
  left_join(sampleMeta, by = "sampleID")

summary(pca)$importance[, 1:6]
```

```{r, fig.width=7, fig.height=5}
ggplot(pcaDf, aes(x = PC1, y = PC2, colour = group, shape = condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = sampleID), vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Control_Static"   = "#6BAED6",
                                 "Control_Laminar"  = "#08519C",
                                 "FAT4 KO_Static"   = "#FD8D3C",
                                 "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = paste0("PC1 (", varExplained[1], "%)"),
       y = paste0("PC2 (", varExplained[2], "%)"),
       title = "PCA of log2 protein abundance") +
  theme_bw()
```

```{r, fig.width=7, fig.height=5}
ggplot(pcaDf, aes(x = PC3, y = PC4, colour = group, shape = condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = sampleID), vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Control_Static"   = "#6BAED6",
                                 "Control_Laminar"  = "#08519C",
                                 "FAT4 KO_Static"   = "#FD8D3C",
                                 "FAT4 KO_Laminar"  = "#A63603")) +
  labs(x = paste0("PC3 (", varExplained[3], "%)"),
       y = paste0("PC4 (", varExplained[4], "%)"),
       title = "PCA of log2 protein abundance") +
  theme_bw()
```

# Session information

```{r}
sessionInfo()
```
